Best Practices for Multiple Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-06-05 paper/manuscript.tex
overleaf review round

\paragraph{Perturbation maps}
For RBFE campaigns with more than two ligands in the input series a \textit{perturbation map} or \textit{-network} must be generated: this is simply a set of data that describe which transformations will be executed between the ligands (referred to as \textit{edges}). ~\cite{liu2013lead} Heuristics have shown the more connected the perturbation network the higher the free energy prediction accuracy versus experimental measures. Running a fully-connected network (i.e. simulate transformations between \textit{all} ligands in the campaign) is typically considered too computationally expensive and may not be suited optimally with current estimators such as MLE (\ref{subsec:uncertainty}).~\cite{scheen22data} Several styles of network designs have been proposed in recent years, although no single method seems to consistently outperform all others.~\cite{ries2024kartograf} There exist several methods for optimization of network structure while minimizing the number of perturbations that need to be computed, reducing the resulting computational cost at a minimal decrease in accuracy ~\cite{yang2020optimal,xu2019optimal}. Sometimes - depending on the RBFE implementation - the introduction of intermediate ligands that are not part of the original congeneric series can be helpful for avoiding challenging transformations such as ring breaking or transformations that would otherwise result in large numbers of atoms being inserted or deleted.~\cite{schoenmaker2025imerge} Some commercial tools have well-calibrated underlying heuristics to plan these network designs but may fail with complicated input datasets, thus requiring manual user validation, in particular when dealing with chiral compounds.

@@

In some cases, when dealing with for example very large datasets that would benefit from rougher initial free energy ranking or in cases where perturbations would be rather large, a star shaped network as seen in Fig.~\ref{fig:fig_types_of_networks} \textbf{A} is used. However, adding redundancy into the network means that a better error analysis can be carried out by analyzing cycle closure errors as discussed in sec.~\ref{sec:are-they-good}, with an example given in Fig.~\ref{fig:fig_types_of_networks}\textbf{B}.

Methods from experimental design have been applied to the construction of perturbation maps. Yang et al.~\cite{yang2020optimal} showed how to optimize the perturbation map by selecting a fixed number of calculations from the pairwise perturbations so that the resulting set of calculations minimize the total variance. Xu~\cite{xu2019optimal} showed how to optimize the perturbation map by allocating different amounts of simulation time to different pairwise perturbations so as to minimize the total variance, given the total simulation time of all the perturbation calculations. Both approaches lead to substantial reduction in the statistical error of the estimated free energies.

>>>>

\paragraph{Perturbation maps}
\tocsectioncomment{Perturbation maps paragraph: @Mary Pitman/ @Jenke Scheen/ OpenFE team to update discussion of perturbation maps?}
Based on the input ligand series, a perturbation map or network can be planned. Recent heuristics have shown the more connected the perturbation network the better. However, there is a way to optimize network structure while minimizing the number of perturbations that need to be computed reducing the resulting computational cost~\cite{yang2020optimal,xu2019optimal}. Sometimes the introduction of intermediates that are not part of the original congeneric series are essential to avoid ring breaking, or to deal with perturbations that would otherwise result in large numbers of atoms being inserted or deleted. Some commercial tools have good underlying heuristics but may fail with complicated input, needing user validation in particular when dealing with chiral compounds.

@@

In some cases, during the lead optimization stage, or for very large datasets that would benefit from rougher initial free energy ranking, or in cases where perturbations would be rather large, a star shaped network as seen in Fig.~\ref{fig:fig_types_of_networks} \textbf{A} is used. However, adding redundancy into the network means that a better error analysis can be carried out by looking at cycle closure errors as discussed in sec.~\ref{sec:are-they-good}, with an example given in Fig.~\ref{fig:fig_types_of_networks}\textbf{B}.

Methods from experimental design have been applied to the construction of the perturbation maps. Yang et al.~\cite{yang2020optimal} showed how to optimize the perturbation map by selecting a fixed number of calculations from the pairwise perturbations so that the resulting set of calculations minimize the total variance. Xu~\cite{xu2019optimal} showed how to optimize the perturbation map by allocating different amounts of simulation time to different pairwise perturbations so as to minimize the total variance, given the total simulation time of all the perturbation calculations. Both approaches lead to substantial reduction in the statistical error of the estimated free energies.

{\color{blue}
\paragraph{Multistate methods}
When dealing with large datasets, methods that compute the binding free energies of multiple ligands simultaneously offer a compelling alternative to pairwise calculations. These approaches -- categorized as "multiple end state" or "multistate" free energy calculations -- sidestep ligand network construction entirely and provide all pairwise free energy differences directly with similar statistical uncertainties (i.e., cycle closure is given by definition). One potential disadvantage of multistate methods is, however, that the presence of multiple end states may complicate the energy landscape, thus requiring longer, inherently sequential simulations. Additionally, they are often more difficult to set up as the combination of multiple ligand topologies into a dual or hybrid topology raises extra challenges \cite{ries2022restraintmaker}.

A well-established method from this category is enveloping distribution sampling (EDS) \cite{christ2007enveloping}, which constructs a single reference state $R$ from the end-states, 
\begin{equation}\label{eq:eds}
H_R (\vec{q}) = - \beta s \ln{ \sum_{i=1}^N \left( e^{-\beta s (H_i(\vec{q})-E_i^R)}\right)},
\end{equation}
where $\vec{q}$ are the coordinates of the system, $\beta=(k_B T)^{-1}$ with $k_B$ being the Boltzmann constant and $T$ the absolute temperature, $N$ is the number of end-states, $s$ is the smoothing parameter and $E^R$ the energy offsets.
The computational efficiency of EDS is mainly due to the fact that interactions between unperturbed protein and solvent degrees of freedom have to be calculated only once regardless of $N$, provided a pairwise decomposable method like e.g. reaction field \cite{tironi1995generalized} is used for the long-range electrostatic interactions. Thus, the computational cost of adding more ligands grows approximately linearly rather than quadratically (in a fully connected perturbation graph) \cite{champion2023accelerating}.

To ensure sampling of all end-states in the EDS simulation, the reference-state parameters $s$ and $E^R$ must be optimized.
To simplify the calibration when $N >2$ and to improve the robustness of the methodology, multiple EDS reference-state potentials can be combined in a replica-exchange setup (termed RE-EDS) with different magnitudes of smoothing ($s$-parameter) \cite{sidler2016replica, sidler2017efficient, ries2022relative, champion2023accelerating}. 
RE-EDS simulations can be conducted using either a single or dual coordinate framework \cite{ries2022restraintmaker}. In the first case, all ligands must be restrained together (e.g. with distance restraints) to avoid diffusing out of the binding site. However, the number and definition of restraints should be chosen carefully as they can introduce bias in the simulations if not explicitly corrected. Single coordinate topologies (such as SCDT) can alleviate this issue as well as improve phase-space overlap \cite{champion2023accelerating} although their implementation in most MD engines usually requires custom routines.

RE-EDS implementations in the GROMOS \cite{champion2023accelerating} or OpenMM \cite{rieder2022leveraging} software packages include automated parameter calibration procedures (i.e., optimizing the distribution of replicas in smoothness-parameter space, optimizing the energy offsets). Previous work showed that 15-30 ligands can be treated in a RE-EDS simulations in practice without issues \cite{rieder2022leveraging, champion2023accelerating,champion2024multistate}.
However, sampling issues can occur for ligands that require insertion or deletion of many heavy atoms. In these cases, it is beneficial to split the set of ligands into smaller groups or include such a ligand via pairwise calculations.
}
